---
title: "Homework Assignment 2"
author: ""
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

## March work day (10 pts)

R functions for common set operations (when you have *actual sets/vectors* of outcomes):

```{r, echo = TRUE, eval = FALSE}
x <- c(1, 2, 3)
y <- c(2, 3, 4, 5)
union(x, y)      # combined elements
intersect(x, y)  # shared elements
setdiff(x, y)    # elements in x but not y (order matters)
setequal(x, y)   # are the sets equal?
```

You have a vector of March 2026 dates (`march`), and vectors `weekends` and `springbreak`. Use R’s set operations to find the dates of **non‑vacation weekdays**.

```{r}
march <- 1:31
weekends <- c(1, 7, 8, 14, 15, 21, 22, 28, 29)
springbreak <- 16:20

workdays <- setdiff(march, weekends)
workdays <- setdiff(workdays, springbreak)
workdays

workdays2 <- setdiff(march, union(weekends, springbreak))
setequal(workdays, workdays2)
```

## Two yellow balloons (20 pts)

**Problem :**  
There are 16 balloons: 5 yellow, 8 red, 3 green. Two darts pop two balloons (without replacement).

Let:

- $P(1_{yellow})$ = probability first balloon is yellow
- $P(2_{yellow})$ = probability second balloon is yellow

1. Given first is yellow, find $P(2_{yellow}\mid 1_{yellow})$.

```{r}
# Given
total <- 16
yellow <- 5

# After popping one yellow balloon, there are 15 balloons left, 4 of which are yellow
remaining_total <- total - 1
remaining_yellow <- yellow - 1

p_2yellow_given_1yellow <- remaining_yellow / remaining_total
p_2yellow_given_1yellow
```


2. Find $P(1_{yellow}\cap 2_{yellow})$ (both balloons yellow).

```{r}
# Multiplication rule: P(A ∩ B) = P(A) * P(B | A)
p_1yellow <- yellow / total
p_both_yellow <- p_1yellow * p_2yellow_given_1yellow
p_both_yellow
```


## Sequential system with A, B, C (15 pts)

A sequential system fails if **any** component fails. Components fail independently with:

- $p_A=0.02$
- $p_B=0.01$
- $p_C=0.03$

Find $P(\text{system fails})$.

```{r}
# Component failure probabilities
p_A <- 0.02
p_B <- 0.01
p_C <- 0.03

# System works only if ALL components work
# P(system works) = P(A works) * P(B works) * P(C works)  (independence)
p_system_works <- (1 - p_A) * (1 - p_B) * (1 - p_C)

# P(system fails) = 1 - P(system works)
p_system_fails <- 1 - p_system_works
p_system_fails
```

## Problem parallel system (15 pts)

A parallel system functions if **at least one** component functions. Components function **independently** with:

- $p_D=0.78$
- $p_E=0.75$
- $p_F=0.82$

Find $P(\text{parallel system functions})$.

```{r}
# Component functioning probabilities
p_D <- 0.78
p_E <- 0.75
p_F <- 0.82

# Parallel system fails only if ALL components fail
# P(all fail) = P(D fails) * P(E fails) * P(F fails)  (independence)
p_all_fail <- (1 - p_D) * (1 - p_E) * (1 - p_F)

# P(system functions) = 1 - P(all fail)
p_parallel_works <- 1 - p_all_fail
p_parallel_works
```

## Spam detection using keywords problem (Bayes' Theorem) (20 pts)


An email service provider is designing a spam filter. From historical data, the company knows:

* **20%** of all incoming emails are spam.
* A specific keyword (for example, “FREE!!!”) appears in:

  * **60%** of spam emails
  * **5%** of legitimate (non-spam) emails

Suppose a new email arrives and it **contains this keyword**. What is the probability that the email is spam?

```{r}
# Given
p_spam <- 0.20
p_legit <- 1 - p_spam  # 0.80

p_keyword_given_spam <- 0.60
p_keyword_given_legit <- 0.05

# Step 1: Total probability of seeing the keyword (law of total probability)
# P(keyword) = P(keyword | spam) * P(spam) + P(keyword | legit) * P(legit)
p_keyword <- p_keyword_given_spam * p_spam + p_keyword_given_legit * p_legit
p_keyword

# Step 2: Bayes' Theorem
# P(spam | keyword) = P(keyword | spam) * P(spam) / P(keyword)
p_spam_given_keyword <- (p_keyword_given_spam * p_spam) / p_keyword
p_spam_given_keyword
```

## Two way contingency table (20 pts)

Suppose you are helping the police develop a new test for blood alcohol levels. The potential advantage of the test is that it can be used during a routine traffic stop, but neither you nor the police are sure how accurate the test is. By design, the test should be positive when blood alcohol level is above $0.05\%.$

The test has been used on a large number of subjects where a more expensive, less convenient test is known to give perfectly accurate results. The table below shows the proportions for the outcome of the new test and true blood alcohol status:

| Test result | Alcohol level > 0.05%  | Alcohol level <= 0.05% |
|-------------|------------------------|------------------------|
| Positive    | 0.08                   | 0.3                    |
| Negative    | 0.02                   | 0.6                    |
	
a) What is the probability that a randomly selected driver has both a positive test and a blood alcohol level higher than $0.05\%$?

```{r}
# Joint probabilities from the table
p_pos_high <- 0.08
p_pos_low  <- 0.30
p_neg_high <- 0.02
p_neg_low  <- 0.60

# a) P(positive AND alcohol > 0.05%): given
p_pos_high
```

b) For a randomly selected driver, what is the probability that the test will be positive?

```{r}
# b) P(positive) = P(pos AND high) + P(pos AND low)
p_pos <- p_pos_high + p_pos_low
p_pos
```

c) Among drivers who have a positive test, what is the probability that a driver has a blood alcohol level higher than $0.05\%$?

```{r}
# c) P(high | positive) = P(pos AND high) / P(pos)
p_high_given_pos <- p_pos_high / p_pos
p_high_given_pos
```

d) Suppose $A$ is the event that a driver has a positive test result and $B$ is the event that the driver has a blood alcohol level above $0.05\%$. Are $A$ and $B$ independent? Justify your answer.

```{r}
# d) A and B are independent if P(A $\cap$ B) = P(A) * P(B)
p_high <- p_pos_high + p_neg_high      # P(B) = P(alcohol > 0.05%)

p_A_and_B <- p_pos_high                # P(A $\cap$ B) = 0.08
p_A_times_B <- p_pos * p_high          # P(A) * P(B)

p_A_and_B
p_A_times_B
```

Since P(A $\cap$ B) $\neq$ P(A) * P(B), A and B are not independent

