---
title: "HW3_Instruction"
author: "Colin Jamison"
format: pdf
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# HW 03- Introduction (100 pts).
## Assignment Overview
The purpose of this assignment is to review probability distribution functions, calculating expected value and variance for a random variable.  

## Instructions
- You will answer the questions directly in the Rmarkdown template file provided. 
- Please change your name in the template file.
- How to insert R code chunks: You can quickly insert chunks into your R Markdown file with the keyboard shortcut Cmd + Option + I (Windows Ctrl + Alt + I).
- You can knit a PDF and upload the final pdf file to Canvas.


## Part I Discrete Random Variable Distribution (Total 20 pts, 5 pts each)

Let $X$ be a random variable with the following probability mass function:

\begin{center}
\begin{tabular}{c|cccc}
$X = x$ & 0 & 1 & 2 & 3 \\\hline
$P(X=x)$ & 0.10 & 0.20 & 0.30 & 0.40\\
\end{tabular}
\end{center}

a) Find $P(X\geq 2)$.

$P(X \geq 2) = P(X=2) + P(X=3) = 0.30 + 0.40 = 0.70$

```{r}
x <- 0:3
p <- c(0.10, 0.20, 0.30, 0.40)

sum(p[x >= 2])
```

b) Find $P(X \geq 2|X\geq 1)$.

$$
P(X \geq 2 \mid X \geq 1) = \frac{P(X \geq 2 \cap X \geq 1)}{P(X \geq 1)} 
 = \frac{P(X \geq 2)}{P(X \geq 1)}$$

```{r}
sum(p[x >= 2]) / sum(p[x >= 1])
```

c) Find $E(X)$.

$$E(X) = \sum x \cdot P(X=x)$$

```{r}
ex <- sum(x * p)
ex
```

d) Find $\text{Var(X)}$.

$$\text{Var}(X) = E(X^2) - [E(X)]^2 = \sum x^2 \cdot P(X=x) - \left[\sum x \cdot P(X=x)\right]^2$$

```{r}
sum(x^2 * p) - ex^2
```

## Part II: Expected value and variance (20 pts)

Let $X$ be a random variable that can take two possible values 0 and 1. It has the following probability mass function: P(X=1) = p and P(X = 0) = 1-p. 

\begin{center}
\begin{tabular}{c|cccc}
$X = x$ & 0 & 1 \\\hline
$P(X=x)$ & 1-p & p \\
\end{tabular}
\end{center}

a) Find $E(X)$.

$$E(X) = \sum x \cdot P(X=x) = 0 \cdot (1-p) + 1 \cdot p = p$$

b) Find $\text{Var(X)}$.

$$\text{Var}(X) = E(X^2) - [E(X)]^2$$

Since $X$ only takes values 0 and 1, $X^2 = X$, so $E(X^2) = E(X) = p$.

$$\text{Var}(X) = p - p^2 = p(1-p)$$

## Part III:  Binomial Distribution (Total 40 pts, 10 pts each)
According to data from the CDC, about 37.1\% of adults (individuals 18 years of age or older) in the United States and 57.9\% of children (individuals between 6 months and 17 years of age) in the United States received a flu vaccine during the 2017-2018 flu season. 

a) Consider a random sample of 50 adults.

Let $X \sim \text{Bin}(n=50, p=0.371)$.

    i. Calculate the probability that exactly 20 adults received a flu vaccine.

$$P(X = 20) = \binom{50}{20} p^{20}(1-p)^{30}$$

```{r}
dbinom(20, size = 50, prob = 0.371)
```

    ii. Calculate the probability that exactly 30 adults did not receive a flu vaccine.

Exactly 30 did *not* receive the vaccine means exactly $50 - 30 = 20$ *did*, so $P(X = 20)$, same as part (i).

```{r}
dbinom(20, size = 50, prob = 0.371)
```

b) Consider a random sample of 20 children.

Let $Y \sim \text{Bin}(n=20, p=0.579)$.

    i. What is the probability that at most 10 children received a flu vaccine?

$$P(Y \leq 10) = \sum_{y=0}^{10} \binom{20}{y} p^y (1-p)^{20-y}$$

```{r}
pbinom(10, size = 20, prob = 0.579)
```

    ii. What is the probability that at least 11 children received a flu vaccine?

$$P(Y \geq 11) = 1 - P(Y \leq 10)$$

```{r}
1 - pbinom(10, size = 20, prob = 0.579)
```

c) State two assumptions you needed to make in order to answer parts a) and b). Briefly comment on the extent to which those assumptions were reasonable.

   1. Independence: Each individual's vaccination status is independent of every other individual's. This is reasonable for a random sample, though in practice household members may share vaccination decisions.

   2. Constant probability of success: Each adult has the same probability $p = 0.371$ of being vaccinated (and each child $p = 0.579$). This is reasonable as an average across the population, though in reality the probability varies by region, socioeconomic status, etc.

d) Consider a random sample of $n = 70$ individuals, which consists of 50 adults and 20 children. Let $Z$ represent the total number of individuals who received the flu vaccine in the sample of 70 individuals. Does $Z$ follow a binomial distribution? Explain your answer.

   No. A binomial distribution requires that each trial has the same probability of success. Here, adults have $p = 0.371$ and children have $p = 0.579$, so the probability of vaccination is not constant across all 70 individuals. Therefore $Z$ does not follow a binomial distribution. ($Z$ is the sum of two independent binomials with different success probabilities.)


## Part IV Consider a senior Statistics concentrator with a packed extracurricular schedule, taking five classes, and writing a thesis. Each time she takes an exam, she either scores very well (at least two standard deviations above the mean) or does not. Her performance on any given exam depends on whether she is operating on a reasonable amount of sleep the night before (more than 7 hours), relatively little sleep (between 4 - 7 hours, inclusive), or practically no sleep (less than 4 hours). (20 pts)

When she has had practically no sleep, she scores very well about 30\% of the time. When she has had relatively little sleep, she scores very well 40\% of the time. When she has had a reasonable amount of sleep, she scores very well 42\% of the time. Over the course of a semester, she has a reasonable amount of sleep 50\% of nights, and practically no sleep 30\% of nights.

  a) What is her overall probability of scoring very well on an exam?

     \begin{center}
     \begin{tabular}{c|c|c}
     Sleep Level & $P(\text{Sleep})$ & $P(W | \text{Sleep})$ \\\hline
     $R$ (reasonable) & 0.50 & 0.42 \\
     $N$ (no sleep) & 0.30 & 0.30 \\
     $L$ (little sleep) & $1 - (P(R) + P(N)) = 0.20$ & 0.40 \\
     \end{tabular}
     \end{center}

     By the law of total probability:

     $$P(W) = P(W|N)P(N) + P(W|L)P(L) + P(W|R)P(R)$$

```{r}
p_w <- 0.30 * 0.30 + 0.40 * 0.20 + 0.42 * 0.50
p_w
```

  b) What is the probability she had practically no sleep the night before an exam where she scored very well?

     By Bayes' theorem:

     $$P(N|W) = \frac{P(W|N) \cdot P(N)}{P(W)}$$

```{r}
(0.30 * 0.30) / p_w
```

  c) Suppose that one day she has three exams scheduled. What is the probability that she scores very well on exactly two of the exams, under the assumption that her performance on each exam is independent of her performance on another exam? (10 pts)

     Using $P(W) = 0.38$ from part (a), and independence:

     $$P(\text{exactly 2 of 3}) = \binom{3}{2} P(W)^2 (1 - P(W))^1$$

```{r}
dbinom(2, size = 3, prob = p_w)
```

  d) What is the probability that she had practically no sleep the night prior to a day when she scored very well on exactly two out of three exams?

     Let $A$ = exactly 2 of 3 well. She has one sleep level for the entire night, which determines the success probability for all 3 exams. By Bayes' theorem:

     $$P(N|A) = \frac{P(A|N) \cdot P(N)}{P(A)}$$

     where $P(A)$ is found via total probability over sleep categories:

     $$P(A) = P(A|N)P(N) + P(A|L)P(L) + P(A|R)P(R)$$

```{r}
p_a_n <- dbinom(2, size = 3, prob = 0.30)
p_a_l <- dbinom(2, size = 3, prob = 0.40)
p_a_r <- dbinom(2, size = 3, prob = 0.42)

p_a <- p_a_n * 0.30 + p_a_l * 0.20 + p_a_r * 0.50

(p_a_n * 0.30) / p_a
```




## Part VI Bonus (Optional, the purpose is to encourage your efforts in learning and character growth. Bonus points could be given but not limited to the following efforts)
a. Attend community coding or office hours: please list the date, time and the faculty/student assistant you met.
b. Write a statement of how you overcome the difficulties in life in order to learn.
c. Write your questions regarding lectures. 






