---
title: "Homework Assignment 4"
author: "Colin Jamison"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```

## 1. The Poisson Distribution (40 pts Total)

**Concept:** Models the number of events occurring within a fixed interval of time or space.

### Scenario: Website Traffic

A website receives an average of **5 hits per minute**. Assume that the number of hits the website receives
follows Poisson distribution

1. Find the probability of receiving exactly 3 hits in a minute.

Let $X \sim \text{Poisson}(\lambda = 5)$.

$$P(X = 3) = \frac{e^{-5} \cdot 5^3}{3!}$$

```{r}
dpois(3, lambda = 5)
```

2. Find the probability of receiving less than 10 hits in a minute.

$$P(X < 10) = P(X \leq 9) = \sum_{x=0}^{9} \frac{e^{-5} \cdot 5^x}{x!}$$

```{r}
ppois(9, lambda = 5)
```

3. Simulation & Visualization

a. Generate a vector called hits with length 1000 that simulates the number of hits using rpois().

b. Create a histogram of this data with hist().

c. Add a vertical red line at the mean using abline().

```{r}
hits <- rpois(1000, lambda = 5)
hist(hits, main = "Simulated Website Hits per Minute", xlab = "Number of Hits")
abline(v = mean(hits), col = "red", lwd = 2)
```



## 2. The Normal Distribution (30 pts, 15 pts each)


### Scenario: A factory produces precision ball bearings. The target diameter is 25.00 mm. Due to machine variance, the diameters follow a Normal distribution with a mean ($\mu$) of 25.00 mm and a standard deviation ($\sigma$) of 0.02 mm.

a: Defect Rates — Bearings are considered defective if they are smaller than 24.95 mm or larger than 25.05 mm. Use pnorm() to calculate the total proportion of defective bearings.

Let $X \sim N(\mu = 25.00, \sigma = 0.02)$. A bearing is defective if $X < 24.95$ or $X > 25.05$.

$$P(\text{defective}) = P(X < 24.95) + P(X > 25.05)$$

```{r}
pnorm(24.95, mean = 25.00, sd = 0.02) + pnorm(25.05, mean = 25.00, sd = 0.02, lower.tail = FALSE)
```


b: Identifying Thresholds — The quality control team wants to flag the smallest 1% of bearings for manual inspection. Use qnorm() to find the diameter threshold for this bottom 1%.

We need the value $x$ such that $P(X \leq x) = 0.01$.

```{r}
qnorm(0.01, mean = 25.00, sd = 0.02)
```



---

## 3. The Exponential Distribution (10 pts each Total: 30 pts + Bonus 10 pts)

**Concept:** Models the time between events in a Poisson process.

### Scenario: Call Center Waiting Time

The time between calls at a help desk follows an exponential distribution with a rate  of **2 calls per hour**.
For the help desk, assume arrivals are a Poisson process.

a. Let $T$ be the waiting time (in hours) until the next ticket. What distribution does $T$ follow and what is its rate parameter?

$T$ follows an Exponential distribution with rate parameter $\lambda = 2$, i.e. $T \sim \text{Exp}(\lambda = 2)$.

b. Compute the probability you wait **more than 45 minutes** for the next ticket.

45 minutes = 0.75 hours.

$$P(T > 0.75) = 1 - P(T \leq 0.75) = e^{-\lambda \cdot 0.75}$$

```{r}
pexp(0.75, rate = 2, lower.tail = FALSE)
```

c. Compute the median waiting time.

The median $m$ satisfies $P(T \leq m) = 0.5$, so $m = \frac{\ln(2)}{\lambda}$.

```{r}
qexp(0.5, rate = 2)
```

(Challenge Question) d. Memoryless property check: verify numerically that
   $P(T > s+t \mid T > s) = P(T > t)$ using s = 0.5, t = 0.3.
   
   


