---
title: "Homework 6"
author: "Colin Jamison"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```

## Central Limit Theorem  (30 pts)

1. Simulation study. Recall during class meeting time, we discussed a simulation study on the central limit theorem by experimenting as the sample size $n$ increases, the sampling distribution of the mean $\bar{X}$ will approach a Normal distribution $N(\mu, \sigma^2/n)$ regardless of the distribution of $X$ ($X$ $\sim$ $Bernouli(0.88)$)

In this problem, you are going to

1. Generate random data from Poisson distribution with parameter $\lambda = 1.2$ with sample size of your choice (e.g. n = 10).

2. Compute the sample mean.

3. Repeat 1 and 2 for 1000 times.

4. Plot the distribution of your sample mean, and report the standard deviation of the sample mean.

5. Experiment on different sample sizes (for example, n = 50, 100, 200) and repeat steps 1-4. Please plot the distribution of the sample means, and report the average of the sample means and standard deviation of the sample means in each case.

Let $X \sim \text{Poisson}(\lambda = 1.2)$. By the CLT, $\bar{X} \xrightarrow{d} N(\mu, \sigma^2/n)$ where $\mu = \lambda = 1.2$ and $\sigma^2 = \lambda = 1.2$.

```{r}
simulate_means <- function(n, reps = 1000) {
  replicate(reps, mean(rpois(n, lambda = 1.2)))
}

sample_sizes <- c(10, 50, 100, 200)

par(mfrow = c(2, 2))
for (n in sample_sizes) {
  means <- simulate_means(n)
  hist(means,
       main = paste("n =", n),
       xlab = expression(bar(X)),
       breaks = 30)
  abline(v = 1.2, col = "red", lwd = 2)
}
```

```{r}
# Report mean and SD of sample means for each n
for (n in sample_sizes) {
  means <- simulate_means(n)
  cat("n =", n, "\n",
      " Mean of xbar:", round(mean(means), 4), "\n",
      " SD of xbar:  ", round(sd(means), 4),
      " Theoretical: ", round(sqrt(1.2 / n), 4),
      " Diff: ", round(sqrt(1.2 / n) - sd(means), 5), "\n") 
}
```

As $n$ increases, the distribution of $\bar{X}$ becomes more tightly concentrated around $\mu = 1.2$ and more closely resembles a normal distribution, consistent with the CLT. The observed standard deviation of the sample means approaches $\sqrt{\lambda / n}$.

## Problem 2 (20 pts) Interval Estimation for population proportion

A city council wants to estimate the proportion of residents who support a new tax to fund public parks. They survey a random sample of 400 residents, and 220 of them state they are in favor of the tax. Please calculate a 95% confidence interval for the true proportion of all city residents who support the tax.

```{r}
# Given
n <- 400
x <- 220
```

1. What is a point estimation for the true proportion of all city residents who support the tax?

$$\hat{p} = \frac{x}{n} = \frac{220}{400}$$

```{r}
p_hat <- x / n
p_hat
```

2. Find the critical value based on the confidence level 95%.

For a 95% confidence interval, $\alpha = 0.05$ and $z^* = z_{0.025}$.

```{r}
z_star <- qnorm(0.975)
z_star
```

3. Find the standard error.

$$SE = \sqrt{\frac{\hat{p}(1 - \hat{p})}{n}}$$

```{r}
se <- sqrt(p_hat * (1 - p_hat) / n)
se
```

4. Calculate the margin of error.

$$ME = z^* \cdot SE$$

```{r}
me <- z_star * se
me
```

5. Find the lower bound and upper bound for the interval.

$$\hat{p} \pm ME$$

```{r}
lower <- p_hat - me
upper <- p_hat + me
c(lower, upper)
```

6. Use R function to find the 95\% confidence interval in this case.

```{r}
prop.test(x, n, conf.level = 0.95, correct = FALSE)
```



## Problem 3 (20 pts)

Use simulation method to study the $\chi^2$ distribution.

1. Generate random numbers of size 1000 from standard normal distribution.

2. For each generated number $x_i$, please calculate the square of $x_i$: $y_i = x_i^2$

3. Please plot the histogram of $y$ and comment on the shape of the distribution of $y$.  (Comment: the distribution of $Y$ is considered as Chi square distribution with degree of freedom of 1. )

Let $X \sim N(0, 1)$. Then $Y = X^2 \sim \chi^2(1)$.

```{r}
# 1. Generate 1000 standard normal values
x <- rnorm(1000)

# 2. Square each value
y <- x^2

# 3. Histogram
hist(y, breaks = 50, main = expression(paste("Distribution of ", Y == X^2)),
     xlab = "y", freq = FALSE)
# Overlay theoretical chi-squared(1) density
curve(dchisq(x, df = 1), col = "red", lwd = 2, add = TRUE)
```

The distribution of $Y$ is heavily right-skewed, concentrated near zero with a long right tail. This matches the $\chi^2(1)$ distribution: most squared standard normal values are small, but occasional large $|X|$ values produce large $Y$.


## Problem 4 (30 pts)

According to Pew Research, 62% of Americans think that too little priority was given to meeting the educational needs of K-12 students during the Covid-19 pandemic. Decisions to keep schools closed and not having enough in-person instruction contributed to parents concern about their kids education. A researcher wants to test whether this percentage is less among Democrats. The researcher selects a random sample of 260 Democrats, and finds that 150 of them think that too little priority was given to children's education during the pandemic.

```{r}
# Given
n <- 260
x <- 150
p0 <- 0.62
```

1. What are the appropriate null and alternative hypotheses?

The researcher wants to test whether the proportion among Democrats is *less* than 0.62.

$$H_0: p = 0.62 \quad \text{vs.} \quad H_a: p < 0.62$$

2. What is the point estimation for the population proportion.

$$\hat{p} = \frac{x}{n} = \frac{150}{260}$$

```{r}
p_hat <- x / n
p_hat
```

3. Compute the test statistic

$$z = \frac{\hat{p} - p_0}{\sqrt{\frac{p_0(1 - p_0)}{n}}}$$

```{r}
z <- (p_hat - p0) / sqrt(p0 * (1 - p0) / n)
z
```

4. Find the p value

This is a left-tailed test, so $p\text{-value} = P(Z \leq z)$.

```{r}
p_value <- pnorm(z)
p_value
```

5. Please use appropriate R function to perform the hypothesis test.

```{r}
prop.test(x, n, p = p0, alternative = "less", correct = FALSE)
```




## Bonus Part (Optional, the purpose is to encourage your efforts in learning and character growth. Bonus points could be given but not limited to the following efforts)
a. Attend office hours or community coding: please list the date, time and the faculty/student assistant you met.
b. Write a statement of how you overcome the difficulties in life in order to learn.
c. What questions do you have about the course material on foundations of inference?



